---
title: "Introduction to baycast"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to baycast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
```

# Overview

`baycast` fits a Bayesian Dirichlet-process accelerated failure-time model with factor-analytic spike-and-slab loadings:

- Patients are clustered through a Dirichlet-process mixture on a shared latent factor `U`.
- High-dimensional features are selected via spike-and-slab on the loading matrix `A`.
- Both **point-mass** (default) and **continuous** spike-slab variants are supported through a single argument `sigma_spike2`.

This vignette walks through a small synthetic example.

# Synthetic data

```{r}
library(BayCAST)
set.seed(1)

n <- 100; p <- 200; d_true <- 3

# Latent factors with two clusters
z_true <- sample(1:2, n, replace = TRUE, prob = c(0.6, 0.4))
U <- matrix(rnorm(n * d_true), n, d_true)
U[z_true == 2, ] <- U[z_true == 2, ] + matrix(rep(c(2, -1, 1), each = sum(z_true == 2)), ncol = d_true)

# Sparse loading: only 20 features active
A <- matrix(0, p, d_true)
signal_idx <- 1:20
A[signal_idx, ] <- matrix(rnorm(20 * d_true, sd = 0.8), 20, d_true)
X <- U %*% t(A) + matrix(rnorm(n * p, sd = 0.1), n, p)

# Survival times
W <- cbind(rbinom(n, 1, 0.5))   # one binary covariate
beta_true <- -0.5
log_t <- W * beta_true + ifelse(z_true == 1, 0, 0.8) + rnorm(n, sd = 0.3)
time   <- exp(log_t)
status <- rbinom(n, 1, 0.85)
```

# Fit (point-mass spike-slab)

```{r, results = "hide"}
fit_pm <- baycast(time, status, X, covariates = W,
                 d = 3, K_init = 5, K_max = 20,
                 iters = 1500, burn = 750, thin = 5,
                 verbose = FALSE, seed = 1)
```

# Variable selection

```{r}
sel <- posterior_pip(fit_pm, threshold = 0.5)
cat("Number of selected features:", length(sel), "\n")
cat("True signals recovered:",
    length(intersect(sel, signal_idx)), "/", length(signal_idx), "\n")
```

# Cluster recovery

```{r}
z_hat <- posterior_z(fit_pm)
table(true = z_true, estimated = z_hat)
```

# Covariate effect

```{r}
posterior_beta(fit_pm, cov_names = "treatment")
```

# Diagnostic plots

```{r, fig.width = 7, fig.height = 8}
plot(fit_pm, type = "trace")
```

```{r}
plot(fit_pm, type = "pip", top_pip = 25)
```

# Continuous spike-slab variant

```{r, results = "hide"}
fit_co <- baycast(time, status, X, covariates = W,
                 d = 3, K_init = 5, K_max = 20,
                 iters = 1500, burn = 750, thin = 5,
                 sigma_spike2 = 0.01,
                 verbose = FALSE, seed = 1)
```

```{r}
cat("PM mode:", fit_pm$ssvs_mode,
    " - features selected:", sum(fit_pm$pip > 0.5), "\n")
cat("CO mode:", fit_co$ssvs_mode,
    " - features selected:", sum(fit_co$pip > 0.5), "\n")
```

On sparse synthetic data the two variants typically agree. On dense real-world data they can disagree substantially; see the package paper for a sensitivity analysis.
